---
title: "Initial LTER data exploration"
author: "Julianna Renzi"
date: "3/16/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(tidyverse)
require(here)
require(wesanderson)
```

# Bring in the data

```{r}
# ALGAE
algae <- read.csv(here("LTER_data/MCR_LTER_Annual_Survey_Benthic_Cover_20220311.csv")) %>% 
  # rename sites to match fish data
  mutate(Site = case_when(Site == "LTER 1" ~ "1",
                          Site == "LTER 2" ~ "2",
                          Site == "LTER 3" ~ "3",
                          Site == "LTER 4" ~ "4",
                          Site == "LTER 5" ~ "5",
                          Site == "LTER 6" ~ "6"
                          )) %>% 
  # also get rid of deeper transect 
  filter(Habitat != "Outer 17") %>% 
  mutate(Habitat = case_when(Habitat == "Outer 10" ~ "Forereef",
                          TRUE ~ Habitat
                          )) %>% 
  mutate(Year = as.factor(Year))


# FISHES
fish <- read_csv(here("LTER_data/MCR_LTER_Annual_Fish_Survey_20210108.csv"),
                 # specify column types
                 col_types = cols(.default = "c", # character
                                  Location = "f", # factor
                                  Year = "f",
                                  Site = "f",
                                  Habitat = "f",
                                  Transect = "f",
                                  Swath = "f",
                                  Total_Length = "d", # double
                                  Count = "i", # integer
                                  Biomass = "d",
                                  Coarse_Trophic = "f",
                                  Fine_Trophic = "f")) %>% 
  # get weird of these weird sites, but should figure out what's going on here
  filter(Habitat != "error") %>% 
  # rename habitat to match algal types
  mutate(Habitat = case_when(Habitat == "BA" ~ "Backreef",
                          Habitat == "FO" ~ "Forereef",
                          Habitat == "FR" ~ "Fringing"
                          )) 
  
# Note: we might want to go through and do a little QA/QC for a couple biomass/length entries (for now just not dealing with then)

```

Ultimately we want values for each site + habitat + year in terms of macroalgaal abundance (and diversity?) and fish abundance (and diversity?)

# Fish summaries

Start by getting potentially interesting fish metrics for each site-habitat-year combination

```{r}
fish %>%
  # just get fishes the LTER considers herbivores/detritivores
  filter(Fine_Trophic == "Herbivore/Detritivore") %>% 
  # group by site, habitat, and year
  group_by(Year, Site, Habitat) %>% 
  # get interesting summary statistics
  summarise(Herbiv_abundance = sum(Count, na.rm = TRUE),
            # get biomass as well as abundance
            Herbiv_biomass = sum(Biomass, na.rm = TRUE), # I think this is right but they don't specify for sure that biomass is total biomass and should be summed
            # get species richness for herbivores/detritivores
            Herbiv_spp_rich = n_distinct(Taxonomy)) %>% 
  ungroup() -> all_herbivores


```

We might also want this for big fishes, after reading Shantz et al. 2019 (Ecological Monographs). Try defining "big" herbivores as greater than 20cm (used in the Shantz model)

```{r}
fish %>%
  filter(Total_Length > 20) %>% 
  # just get fishes the LTER considers herbivores/detritivores
  filter(Fine_Trophic == "Herbivore/Detritivore") %>% 
  # group by site, habitat, and year
  group_by(Year, Site, Habitat) %>% 
  # get interesting summary statistics
  summarise(Herbiv_abundance20cm = sum(Count, na.rm = TRUE),
            # get biomass as well as abundance
            Herbiv_biomass20cm = sum(Biomass, na.rm = TRUE), # I think this is right but they don't specify for sure that biomass is total biomass and should be summed
            # get species richness for herbivores/detritivores
            Herbiv_spp_rich20cm = n_distinct(Taxonomy)) %>% 
  ungroup() -> big_herbivores

```


# Now look at algal data

Get broad "coral" cover

```{r}
algae %>% 
  filter(Taxonomy_Substrate_Functional_Group == "Coral") %>% 
  # change site and habitat designations to match
  group_by(Year, Site, Habitat) %>% 
  summarize(Avg_coral_cover = mean(Percent_Cover)) -> coral_cover

```

Now get more specific algal metrics

```{r}
algae %>% 
  # select just the algal entries 
  filter(Taxonomy_Substrate_Functional_Group != "Coral" & 
           Taxonomy_Substrate_Functional_Group != "Sand" & 
           Taxonomy_Substrate_Functional_Group != "Ascidian" & 
           Taxonomy_Substrate_Functional_Group != "Sponge" & 
           Taxonomy_Substrate_Functional_Group != "Millepora platyphylla" & 
           Taxonomy_Substrate_Functional_Group != "Shell Debris" & 
           Taxonomy_Substrate_Functional_Group != "Corallimorpharia" & 
           Taxonomy_Substrate_Functional_Group != "Soft Coral" & 
           Taxonomy_Substrate_Functional_Group != "Cyanophyta" & 
           Taxonomy_Substrate_Functional_Group != "Coral Rubble" & 
           Taxonomy_Substrate_Functional_Group != "Symploca hydnoides" & 
           Taxonomy_Substrate_Functional_Group != "No data" & 
           Taxonomy_Substrate_Functional_Group != "Tridacna sp." & 
           Taxonomy_Substrate_Functional_Group != "Bare Space" & 
           Taxonomy_Substrate_Functional_Group != "Sarcophyton sp."
           ) %>% 
  # make a column for genus
  rowwise() %>% 
  mutate(Genus = str_split(Taxonomy_Substrate_Functional_Group, " ")[[1]][1]) %>%
  # do some clean up
  mutate(Genus = case_when(Genus == "Algal" ~ "Turf", 
                           Genus == "Crustose" ~ "CCA",
                           Genus == "Damselfish" ~ "Turf",
                           TRUE ~ Genus
                           )) -> algae
  
  
 
  
```

## Get metrics for overall algae by site

First for algae

```{r}
algae %>% 
  group_by(Year, Site, Habitat) %>% 
  filter(Genus != "CCA" &
           Genus != "Turf") %>% 
  summarize(Avg_algal_cover_noCCAnoTurf = mean(Percent_Cover),
            Number_algal_genera_noCCAnoTurf = n_distinct(Genus)) %>% 
  ungroup() -> overall_algae
```

Then for turf, CCA, Lobophora, and Turbinaria (of some interest)

```{r}
# for CCA
algae %>% 
  group_by(Year, Site, Habitat) %>% 
  filter(Genus == "CCA") %>% 
  summarize(Avg_CCA_cover = mean(Percent_Cover)) %>% 
  ungroup() -> overall_CCA


# for turf
algae %>% 
  group_by(Year, Site, Habitat) %>% 
  filter(Genus == "Turf") %>% 
  summarize(Avg_turf_cover = mean(Percent_Cover)) %>% 
  ungroup() -> overall_turf

# for Lobophora
algae %>% 
  group_by(Year, Site, Habitat) %>% 
  filter(Genus == "Lobophora") %>% 
  summarize(Avg_lobophora_cover = mean(Percent_Cover)) %>% 
  ungroup() -> overall_lobophora

# for Turb
algae %>% 
  group_by(Year, Site, Habitat) %>% 
  filter(Genus == "Turbinaria") %>% 
  summarize(Avg_turb_cover = mean(Percent_Cover)) %>% 
  ungroup() -> overall_turb


```


## Join data together

```{r}
all_herbivores %>% 
  full_join(big_herbivores, by = c("Year", "Site", "Habitat")) %>% 
  full_join(coral_cover, by = c("Year", "Site", "Habitat")) %>% 
  full_join(overall_algae, by = c("Year", "Site", "Habitat")) %>% 
  full_join(overall_CCA, by = c("Year", "Site", "Habitat")) %>% 
  full_join(overall_turf, by = c("Year", "Site", "Habitat")) %>% 
  full_join(overall_lobophora, by = c("Year", "Site", "Habitat")) %>% 
  full_join(overall_turb, by = c("Year", "Site", "Habitat")) %>% 
  # remove years that we don't have fish data for yet
  filter(Year != 2005,
         Year != 2006, # also getting rid of 2006 because algal surveys were split over 2005 & 2006 and will need to make some judgment calls for how to deal with this
         Year != 2020,
         Year != 2021) %>% 
  # then fill in missing values with 0 (e.g., no coral/algae/turf observed)
  replace(is.na(.), 0) -> fishNalg

```

# Plot it up

Bring in colors

```{r}
# for forereef/fringing/backreef
HabitatPal <- wes_palette(name = "BottleRocket2", n = 3)
```

## Herbivores v. Coral

Plot just overall herbivore biomass

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_coral_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average coral cover (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")

```

Then think about "bigger" herbivores

- Not sure why my filter isn't working here--need to come back to that

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass20cm, y = Avg_coral_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average coral cover (%)") +
  xlab("Total big herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Biomass of big herbivores (>20cm) by LTER site")


```


## Herbivores v. algae

Plot just overall herbivore biomass

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_algal_cover_noCCAnoTurf, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average algal cover, no CCA or turf (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")
```

CCA

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_CCA_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average CCA cover (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")
```


Turf

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_turf_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average turf cover (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")
```


Turbinaria

- This looks almost unimodal.... Can't think of why that would be?

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_turb_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average Turbinaria cover (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")

```

Lobophora

```{r}
fishNalg %>% 
  ggplot(aes(x = Herbiv_biomass, y = Avg_lobophora_cover, color = Habitat)) +
  geom_point() +
  facet_wrap(.~Site) +
  scale_color_manual(values = HabitatPal) +
  ylab("Average Lobophora cover (%)") +
  xlab("Total herbivore biomass (g)") +
  theme_bw() +
  ggtitle("Total herbivore biomass by LTER site")

```




